---
title: "Exercise 4. Risk classification"
subtitle: "Report"
author: "Your name"
date: today
format: 
  html:
    toc: true
    code-fold: true
    echo: false
    message: false
    warning: false
editor: visual
---

```{r}
library(readr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(pROC)
```

## Data visualisation and initial comparison of predictors

Which of the two predictors do you think is the better indicator to classify cancers into malignant or benign?

Answer:
<!-- TYPE YOUR ANSWER HERE -->

```{r}
#| warning: false

## Load data
df <- as_tibble(read_csv("../data/breast-cancer.csv"))
```

```{r, out.width = "50%"}
df %>% select(c(diagnosis, radius_mean, compactness_mean)) %>% 
ggplot(aes(x=radius_mean, fill=diagnosis)) +
  geom_density(alpha=0.5)

df %>% select(c(diagnosis, radius_mean, compactness_mean)) %>% 
ggplot(aes(x=compactness_mean, fill=diagnosis)) +
  geom_density(alpha=0.5)
```


## Frequency of errors for the binary classification model with mean radius as predictor and cutoff 15

```{r}
#| output: false

# output from the function table. You can hide the output by typing #| output: false

table(df %>% select(c(diagnosis, radius_mean, compactness_mean)) %>%
  mutate(pred_rad=case_when(radius_mean>15 ~ "pos", .default = 'neg')) %>% 
  select(c(diagnosis,pred_rad)))
```

For the binary classification model using mean radius as predictor:

Frequency of False positives (FP) is  <!-- TYPE YOUR ANSWER HERE -->

Frequency of False negatives (FN) is  <!-- TYPE YOUR ANSWER HERE -->

Which of these two errors do you think is worse? Motivate your answer.

<!-- TYPE YOUR ANSWER HERE -->


## Sensitivity and specificity for the binary classification model with mean radius as predictor and cutoff 15

For the binary classification model using mean radius as predictor:

Frequency of True Positives (TP) is <!-- TYPE YOUR ANSWER HERE -->

Frequency of True Negatives (TN) is <!-- TYPE YOUR ANSWER HERE -->

Sensitivity, i.e. $\frac{TP}{TP + FN}$ is <!-- TYPE YOUR ANSWER HERE -->

Specificity $\frac{TN}{FP + TN}$ is <!-- TYPE YOUR ANSWER HERE -->

## Model comparison using the ROC curve methodology

```{r}
#| message: false
r = roc(diagnosis ~ radius_mean, data=df)
ssc = coords(r, "best", best.method = "closest.topleft")

r2 = roc(diagnosis ~ compactness_mean, data=df)
ssc2 = coords(r2,"best", best.method = "closest.topleft")
```

```{r}
list(radius=r,compactness=r2) %>% ggroc +
  annotate("text", x = 0.3, y = 0.105, label = paste0("AUC radius = ", round(auc(r), 2))) +
  annotate("text", x = 0.3, y = 0.05, label = paste0("AUC compactness = ", round(auc(r2), 2))) +
  ggtitle("ROC curves")
```

Which of the two binary classification models have the best performance according to the AUC measure?

<!-- TYPE YOUR ANSWER HERE -->

Suggest three things that could be done to build a better classification model?

- suggestion 1

- suggestion 2

- suggestion 3
