---
title: Monte Carlo simulation - listeria
author: Ullrika
#echo: false
warning: false
---

```{r}
library(ggplot2)
```

## A challenge in reproducing an assessment

It turns out that this is a good exercise is the challenges that can come from reproducing the work by someone else. It is also a way to learn the importance of documentation. 



## Model

>Ullrika have said - "Don't trust anyone - anyone can be wrong". It also applies to her. 

In this case the model equation in the study material has an error: $EGRT(T\cdot t)$ should be $EGRT(T)\cdot t$

The model consists of two equations: 

$$C(t)=\frac{C_{max}}{1+(C_{max}/C(0)-1) \cdot \exp(-EGR(T)\cdot t)}$$
$$EGR(T)=EGR(5)\left(\frac{T-T_{min}}{5-T_{min}}\right)^2$$
**State variables:** $C(t)$

**Parameters:** $EGRT(5)$ and $C(0)$ which are treated as uncertain. $C_{max}$ and $T_{min}$ which are set to be fixed. 

**Other assessment values:** $t = 2.5$


## Set up and run Monte Carlo simulation

```{r}
niter = 2000
t = 2.5
```

### Fixed parameters

$log10 C_{max} = 7.83$ from table 2 lowest row

```{r}
log10Cmax = 7.83
Cmax = 10^log10Cmax
```

$T_{min} = -1.18$ from the text

```{r}
Tmin = -1.18
```


### Uncertainty in initial concentration

Initial concentration on the log10 cpu/g scale is modelled by a generalised beta distribution in the interval with the lower (L) and upper (U) bounds -1.69 to 7. Parameter values taken from Table C.4 RTE category Cheese. 

$$\frac{C(0)-L}{U-L}= \frac{C(0)-(-1.69)}{7-(-1.69)} \sim beta(0.194,3.177)$$

```{r}
log10C0 <- rbeta(niter, 0.194, 3.177)*(7--1.69)+-1.69 # log10 cpu/g
C0 = 10^log10C0 #transform to original scale

#log10C0 <- rnorm(niter, -1.19, 0.968) # used in class, values comes from Maartens simulation
#EGRT20 <- rnorm(niter,0.137,0.0137) # used in class
```


## Uncertainty in growth at 5 degrees

::: {note}
The lognormal distribution

To say that $Y$ is lognormal distributed implies that the log of $Y$ is normally distributed. 

$$log Y \sim N(\mu,\sigma)$$ 

where $\mu$ and $\sigma$ are the expected value and standard deviation on the log scale. 

Let us denote the expected value and standard deviation $Y$ on its original (real scale) by $m$ and $s$. 

It is common to use the parameters on the log scale to parametrise a lognormal distribution (this is what we do in R). In this case, it is not clear whether the parameters for the lognormal in Table 2 are given on the real and on the log scale. Since they are positive, I suspect they are given on the real scale. 

I also know that the biohaz group that worked with this often used the software @risk to perform Monte Carlo simulation. This software one has two ways of parametrising the lognormal distribution, on the real or on the log scale. 

To transform from real-space to log-space we can do the following 

$$\sigma^2 = \ln (1 + (s/m)^2)$$
$$\mu = \ln(m)-\sigma^2/2$$

In case someone wants to do it the other way around: From log-space to real-space: 

$$m=e^{\mu+\sigma^2/2}$$

$$s = m \cdot \sqrt{e^{\sigma^2}-1}$$

:::

When we compare the two distributions, we see that treating the numbers in table 2 as mean and standard deviation on the real scale generates a reasonable distribution.   

```{r}
#| warning: false

m = 0.0117 # from table 2
s = 0.00816 # from table 2

sigma2 = log10(1+(s/m)^2)
sigma=sqrt(sigma2)
mu=log10(m)-sigma2/2

# draw sample from EGRT5
EGRT5 <- 10^rnorm(niter,mu,sigma) 
EGRT5_logscale <- 10^rnorm(niter,m,s) 

ggplot(data.frame(vals = c(EGRT5,EGRT5_logscale), scale = rep(c("parameters on real scale","parameters on log scale"),each=niter)),
       aes(x=vals,fill=scale)) +
         geom_histogram() +
  facet_wrap(~scale, ncol = 1)
```

## Propagate uncertainty 

```{r}
# transform to EGRT20
EGRT20 =  EGRT5*(20-Tmin)/(5-Tmin)^2 # equation 8
#mean(EGRT20)# I get around 0.09
#EGRT = 0.137 # the point estimate in the paper, he get a higher one. It looks like his distribution has higher values than ours. 


# Output
Ct = Cmax/(1+(Cmax/C0)-1)*exp(-EGRT20*t)
log10Ct = log10(Ct)
```

# Summarise and visualise uncertainty

```{r}
#| eval: false
hist(Ct) # a bit ugly graph
```



```{r}
df_MC <- data.frame(log10Ct = log10Ct) 

ggplot(df_MC, aes(x = log10Ct)) +
  geom_histogram(binwidth = 0.25) +
  geom_vline(xintercept=2,colour = "red")
```


The histogram is a bit edgy. I would go back and increase the number of iterations to make it more smooth. When doing so, I reflect on the value of writing up the MC-simulation as a function.  

## Function for the MC 

```{r}
run_MC <- function(niter){
# parameter uncertainty
  
log10C0 <- rbeta(niter, 0.194, 3.177)*(7--1.69)+-1.69 # log10 cpu/g
C0 = 10^log10C0 #transform to original scale

m = 0.0117 # from table 2
s = 0.00816 # from table 2

sigma2 = log10(1+(s/m)^2)
sigma=sqrt(sigma2)
mu=log10(m)-sigma2/2

# draw sample from EGRT5
EGRT5 <- 10^rnorm(niter,mu,sigma) 
EGRT5_logscale <- 10^rnorm(niter,m,s) 

# propagate 
EGRT20 =  EGRT5*(20-Tmin)/(5-Tmin)^2 # equation 8

Ct = Cmax/(1+(Cmax/C0)-1)*exp(-EGRT20*t)
log10Ct = log10(Ct)
return(data.frame(log10C0=log10C0,EGRT5=EGRT5,EGRT20=EGRT20,log10Ct=log10Ct))
}
```


```{r}
df_104 <- run_MC(10^4)
ggplot(df_104, aes(x = log10Ct)) +
  geom_histogram(binwidth = 0.25) +
  geom_vline(xintercept=2,colour = "red") #+
  #labs(title=paste("P(log10 Ct < 2)=",round(100*mean(df_104$log10C0<2),1),"%")) # a way to add the calculated probability in the graph
```

### Summary that one can provide

```{r}
quantile(log10(Ct),probs=c(0.05,0.95)) # 90% prob int
mean(log10(Ct)) # mean considering uncertainty
```
### Useful summary for decision maker 

```{r}
mean(log10Ct<2)
```
# Answer to the question 

I am `r round(mean(log10Ct<2)*100,1)`\% certain that your cheese is safe.