---
title: Population Viability Analysis
subtitle: MVEN10
author: Ullrika
date: today
format: html
editor: source
warning: false
echo: false
---

```{r}
# load required libraries 

library(ggplot2)
library(dplyr)
```

## Model specification 

$$B_t = B_{t-1} e^{(\pi -\delta B_{t-1}+S_t)} - C_t$$

$$S_t \sim N(0,\sigma)$$


(@) Identify state variables, assessment variables and model parameters. 

## PVA with fixed parameter values

```{r}
# initial or fixed state variables
B0 = 250 # Initial biomass of cod
C = 23 # Catch (fixed fishing pressure the time period)

# parameters
pi = 0.5 # growth rate
delta = 0.002 # mortality rate
sigma = 0.25 # standard deviation for environmental variability

# assessment variables
endT = 30 # Period of simulation - set to 3 generation lengths
time = 0:endT  # Time period is set to three generation lengths
qext = 10 # Quasi-extinction threshold
```

```{r}
# model 
niter = 100 #number of times I want to repeat the simulation
below = rep(0,niter) # counter of quasi extinctions


for(i in 1:niter){
  
B = rep(0,endT +1) #create a vector of zeros
B[1] = B0 #assign the starting biomass value
for(t in 1:endT ){
  B[t+1] = B[t]*exp(pi-delta*B[t]+rnorm(1,0,sigma)) - C
  B[t+1] = max(0,B[t+1])
}

below[i] = max(B < qext) #this take the value one if below the quasi extinction threshold

# plot 
if(i == 1){
plot(time,B, ylab = "biomass", type = "l", ylim = c(0,500))
abline(h = qext, col = "red")
}else{
 lines(time,B) 
}

}
ext_risk = mean(below) # quasi extinction risk i.e. the likelihood for extinction 
mtext(ext_risk,side=3) # adds this information to the plot

```

## Evaluate robustness in the PVA

(@) Find the number of iterations where the estimate of the quasi-extinction risk has a reasonable standard deviation.

```{r}
# create a function of the above generated code to run a simulation given known parameter values and where the number of iterations is an input argument
calc_risk <- function(niter_ale){
#niter_ale is the number of times I want to repeat the simulation to capture variability

# parameters (fixed)
par = data.frame(pi = 0.5, delta = 0.002, sigma = 0.25)

# model 
below = rep(0,niter_ale) #counter of quasi extinction

for(i in 1:niter_ale){
  
B = rep(0,endT +1)
B[1] = B0
for(t in 1:endT ){
  B[t+1] = B[t]*exp(par$pi-par$delta*B[t]+rnorm(1,0,par$sigma)) - C
  B[t+1] = max(0,B[t+1])
}

below[i] = max(B < qext) #one if below threshold

}
ext_risk = mean(below) #quasi extinction risk i.e. the likelihood for extinction 

return(ext_risk)
}
```

## Robustness check

```{r}
#| eval: false

# run the function n=200 times for different choices of niter (10, 50, 100, 500, 1000)
rob_risk <- replicate(n=200,calc_risk(niter_ale=10)) #takes long time to run if you increase any of the two numbers
#summary(rob_risk)
sd(rob_risk)
```



(@) Make a plot showing the standard deviation for different number of iterations 

```{r}
## here is way to loop through the different options and save into a data frame 

df_rob <- do.call('rbind',lapply(c(10, 50, 100, 500, 1000),function(j){
  rob_risk <- replicate(n=200,calc_risk(niter_ale=j)) 
  data.frame(niter_ale = j, sd = sd(rob_risk))
}))

ggplot(df_rob, aes(x=niter_ale, y=sd)) +
  geom_line()
```

(@) Make a plot showing the estimated quasi-extinction risk with error bars for different number of iterations. Add "quasi-extinction risk" as y-label. Add "number of iterations for variability" as x-label.

```{r}
## here is way to loop through the different options and save into a data frame 

df_rob2 <- do.call('rbind',lapply(c(10, 50, 100, 200, 500),function(j){
  rob_risk <- replicate(n=200,calc_risk(niter_ale=j)) 
  data.frame(niter_ale = j, sd = sd(rob_risk), m = mean(rob_risk))
}))

ggplot(df_rob2, aes(x=niter_ale, y=m)) +
  geom_point(colour = "brown") +
  geom_errorbar(aes(x=niter_ale,ymin=m-sd,ymax=m+sd)) +
  ylab("quasi-extinction risk") +
  xlab("number of iterations for variability")
```

## Consider uncertainty in parameters 

So far we used point estimates of the parameters. Now we will include uncertainty in parameters. Since we have both uncertainty and variability, this will be a 2D Monte Carlo simulation. 

```{r}

# Draw random numbers to represent uncertainty in parameters

niter_epi = 200 # number of random draws. we can change this later on

# draws saved into a data frame
df_MC <- data.frame(
  pi = rnorm(n = niter_epi, mean = 0.5, sd = 0.05),
  delta = rnorm(n = niter_epi, mean = 0.002, sd = 0.0002),
  sigma = abs(rnorm(n = niter_epi, mean = 0.25, sd = 0.1)) # put absolute to avoid getting negative values which are not possible for a standard deviation
  )


```

Here I have copied and pasted in the function for the assessment, I use the selected number of iterations for variability from above (niter_ale) and have given it a different name (calc_risk_2D). The input argument is now iter_epi. 

(@) Add uncertainty in the parameters into the Monte Carlo simulation of the PVA. Replace parameter values by the iter_epi'th draw in df_par. 

```{r}

calc_risk_2D <- function(iter_epi){
  
niter_ale = 500 # put in your selected value here

# parameters
#par = data.frame(pi = 0.5, delta = 0.002, sigma = 0.25) # replace this
par = df_MC[iter_epi,]

# model 
below = rep(0,niter_ale) #counter of quasi extinction


for(i in 1:niter_ale){
  
B = rep(0,endT +1)
B[1] = B0
for(t in 1:endT ){
  B[t+1] = B[t]*exp(par$pi-par$delta*B[t]+rnorm(1,0,par$sigma)) - C
  B[t+1] = max(0,B[t+1])
}

below[i] = max(B < qext) #one if below threshold


}
ext_risk = mean(below) #quasi extinction risk i.e. the likelihood for extinction 

return(ext_risk)
}
```

(@) Run the PVA simulation for all niter_epi draws from parameter uncertainty and save to an object

```{r}
unc_risk <- sapply(1:niter_epi,calc_risk_2D)
unc_risk
```

(@) Visualise uncertainty in the risk estimate using a histogram 

```{r}
df_MC$risk <- unc_risk # saving the output in the same dataframe as unc in parameters

ggplot(df_MC,aes(x=risk)) +
  geom_histogram(binwidth = 0.05) +
  geom_vline(xintercept=0.10, col = "red") +
  xlim(NA,1)
```

(@) Visualise uncertainty in the risk estimate using a boxplot. Add "Uncertainty in quasi-extinction risk" as a title. Add "Quasi extinction risk" as y-label. Make y axis go from 0 to 1. Flip axis.

```{r}
df_MC$C <- as.character(C) # add the value of the catch variable in the data frame

ggplot(df_MC,aes(x=C, y=risk)) +
  geom_boxplot() +
  labs(title="Uncertainty in quasi-extinction risk")+
  ylab("Quasi extinction risk") +
  ylim(0,1) +
  coord_flip() 
  
```

(@) Summarise uncertainty in the quasi-extinction risk by a 90% probability interval and a mean value. 

```{r}
probint = quantile(df_MC$risk,probs=c(0.05,0.95))
m = mean(df_MC$risk)
```

(@) Add the results into the following text.  

The quasi-extinction risk for a catch of `r df_MC$C[1]` is assessed to be `r m` with an 90% uncertainty interval (`r probint[1]` , `r probint[2]`). 

## Sensitivity analysis 

One way to identify which parameters that has the strongest contribution to uncertainty in the output is to perform a sensitivity analysis. I recommend to perform sensitivity analysis by visualisatio and a quantitative measure. 

(@) Visualise the influence of parameter on calculated risk using scatter plots

```{r}
plot(df_MC[,c("pi","delta","sigma","risk")])
```

(@) Calculate the spearman rang correlation 

```{r}
cor(df_MC[,c("pi","delta","sigma","risk")], method = "spearman")
```

(@) Why is the influence from growth negative? 

(@) Which parameter has the strongest influence on the calculated risk? 

Which parameter that has the strongest influence depends on the role it has in a model and how much uncertainty it has compared to the other parameters. 

## Decision analysis

Decision makers wants to find the fishing pressure for which the experts are at least 80\% certain that the quasi-extinction risk is below 10. 

We search for the highest fishing pressure where this applies, but testing different values on C and summarising uncertainty by its 80\% quantile. 

```{r}
df_DA <- do.call('rbind',lapply(c(10,12,14,15,16,17,18,20),function(c_val){
C = c_val

calc_risk_2D <- function(iter_epi){
  
niter_ale = 500 # our selected value where we consider the estimate is reasonably robust with respect to variability

# parameters
#par = data.frame(pi = 0.5, delta = 0.002, sigma = 0.25) # replace this
par = df_MC[iter_epi,]

# model 
below = rep(0,niter_ale) #counter of quasi extinction


for(i in 1:niter_ale){
  
B = rep(0,endT +1)
B[1] = B0
for(t in 1:endT ){
  B[t+1] = B[t]*exp(par$pi-par$delta*B[t]+rnorm(1,0,par$sigma)) - C
  B[t+1] = max(0,B[t+1])
}

below[i] = max(B < qext) #one if below threshold


}
ext_risk = mean(below) #quasi extinction risk i.e. the likelihood for extinction 

return(ext_risk)
}

unc_risk <- sapply(1:niter_epi,calc_risk_2D)
df_MC$risk <- unc_risk # saving the output in the same dataframe as unc in parameters
df_MC$C <- as.character(c_val) # to make this into a non-numeric entry
return(df_MC)
})
)
```

Visualise the simulated values. 

```{r}
ggplot(df_DA,aes(x=C, y=risk)) +
  geom_boxplot() +
  ylab("Quasi-extinction risk") +
  xlab("Catch") +
  labs(title="Uncertainty in the quantity of interest") +
  geom_hline(yintercept=0.1,colour="red")
```

### Summarise uncertainty for different catch values 

```{r}
df_DA %>% #for this you need dplyr
  group_by(C) %>%
  summarise(av_risk=mean(risk),sd_risk = sd(risk),q80=quantile(risk,probs=0.8))
```
